---
title: "Error catching"
editor: visual
---

<br><br>

Two options worth considering, `tryCatch` and `withCallingHandlers`. 

This distinction trips people up. tryCatch() exits the protected code when a condition is caught — you can't go back. withCallingHandlers() handles the condition in place and, for restartable conditions like warnings and messages, execution can continue afterward.  

You may also like `purrr::possibly` and `purrr::safely`.

## `tryCatch`  

`tryCatch` key arguments:  

1. `expr =`, the code you're 'protecting'.   
2. `error =`, where an error is returned it runs the expression in the error curly brackets.  
3. `warning =`, likewise, the warning evaluates the result in its curly bracket.  

Whatever the output of the matched condition is what `tryCatch` returns. 

`finally` always runs, but if it returns code that is always discarded, it's just a printing tool, or a tool for resetting values or closing connections. 

```{r, collapse=T, error=T}

1 / "a"

result <- tryCatch(

  {1 / "a"}, # code that might throw a condition

  error = function(e) {
    message("Caught an error: ", 
            conditionMessage(e), # R's error message
            " from ", 
            deparse(conditionCall(e))) # The code you ran
    NA
  },
  warning = function(w) {
    message("Caught a warning: ", conditionMessage(w))
    NULL
  },
  finally = {
    message("This always runs, success or failure")
  }
)

result # Here 'result' returns NA, the return value of the error curly brackets. 

```

<br>

Using finally to reset values. 

```{r, collapse = T}

result <- tryCatch(
  
  {x = 1
  while(x < 10){
    x = x + 1
  }
  x}, 
  
  finally = {message("x resets to 1")
    x = 1
    }
  
  
)

x
result

```

<br>

When `warning = ` is specified, a warning will stop your code. 

```{r, collapse = T}

result <- tryCatch(
  
  {as.numeric(c("1", "2", "d"))},
  
  warning = function(w){
  message("Regretably, your code has stopped for this warning. \nNow it just returns NA.")
    NA}
  )

result

```

<br>

When both warning and error are requested, the one that is triggered first is what gets returned. 

```{r, collapse=TRUE}

result <- tryCatch(
  {
    warning("a warning")
    stop("an error")
  },
  error = function(e) "Would have caught the error",
  warning = function(w) "But the warning appeared first"
)

result

```

Once an error or warning is specified by tryCatch() this over-rides messages generated by the code itself. 

```{r}

tryCatch({
    warning("a warning")
    })

tryCatch({
    warning("a warning")
    }, 
    warning = function(w){
      "Original warning - over-ridden!"
    })

```

<br>

## purrr::possibly()

`purrr::possibly()` is an error suppression tool for functions.

It takes a function, and defines an object to return when it errors, rather than returning the error. 

```{r, error = TRUE, collapse=T}

safe_log <- 
  purrr::possibly(
    log, # function to make safe
    otherwise = NA_real_ # object to return in place of errors
    )

safe_log(10)

log("a")

safe_log("a")

```

Although, the same effect is obtainable with `tryCatch()`. 

```{r}

safe_log <- function(x) {
  tryCatch(
    log(x),
    error = function(e) NA_real_
  )
}

safe_log("a")

```

<br>

## safely()

This also wraps a function so it does not stop on an error, but instead instead it returns a list containing both the result and the error. 

`safely` also has an `otherwise` argument that can be specified so the value returned when erroring is not NULL.

```{r, collapse=T}
safe_log <- purrr::safely(log)

# Error is NULL because there was not an error
safe_log(10)

# Now, there is an error value, but result is NULL. 
safe_log("a")

```


































