---
title: "Loading packages from GitHub"
format: html
editor: visual
---

## Creating a link

My packages are typically in a subfolder of the repo, so you need to load the sub-folder specifically.

```{r, collapse = T}

# Take the link from GitHub - 
# https://github.com/Joeacrowley/Descriptives/tree/main/dtab

# But some bits need removing...

# Remove 'https://github.com/' from the URL. 
"Joeacrowley/Descriptives/tree/main/dtab"

# Then remove both 'main' and 'tree':
# - 'tree' is GitHub's web-routing convention, not required here and will cause an error. 
# - 'main' is the branch, required only if you want to specify a branch

dtab_link <- "Joeacrowley/Descriptives/dtab"

# If you wanted to specify branch
"Joeacrowley/Descriptives/dtab@main"

# This quiets warnings relating to my Rstudio theme. 
options(cli.ignore_unknown_rstudio_theme = TRUE)

```

<br>

## Installation with `pak`

Preferred: `pak` (faster, better dependency resolution than remotes/devtools, and understands sub-directories natively).

It builds a summary of package dependence and compatability, checks it, then installs.

It works well with:

-   `devtools`: pkg-development toolkit. load_all(), document(), check(), test(), build(). For building/testing your own packages.

`remotes`: lightweight installer for non-CRAN sources — GitHub, GitLab, Bitbucket, local, URL. install_github() etc. devtools used to bundle this; now imports it. Predates pak. Slower, weaker dep resolution, no conflict pre-check — but still fine, widely used.

```{r, collapse=T}

# Install package
pak::pkg_install(dtab_link)
library(dtab)

# Now can call dtab
dtab::make_numeric_summary_table(
  mtcars, 
  outcomes = names(mtcars), 
  statistics = c("min", "mean", "max")
  ) 

```

<br>

`pak` may struggle at work, it needs its own metadata service which can be blocked by corporate proxies. Alternatively, use `remotes::install_github()` or install from a local source.

```{r, eval=F}
pak::pkg_install("local::C:/path/to/Descriptives/dtab")
```

<br>

The below might work better at work. The installation itself should be the same, but lacks some of the `pak` checks.

```{r, message=T}
remotes::install_github("Joeacrowley/Descriptives", subdir = "dtab")
```
