---
title: "Markdown"
format: html
editor: visual
---

## Nesting tabs

The code for the producing nested tabs with a loop is illustrated below. 

Note the lower level header is exactly one level lower than the top header.

Note the `::: {.panel-tabset}` / `:::` system is for .qmd files only, with .Rmd files you simply write `## {.tabset}` after the **higher level** header, and it cuts out after you write another header of that level. 

The code here is visible only because I've duplicated it outside the `{r}` chunk so it unevaluated, usually it is invisible.

```{}

# ### Level 3 Header
# 
# ::: {.panel-tabset}
# 
# ```{r, results='asis'}
# # Example using mtcars dataset grouped by cylinder count
# groups <- split(mtcars, mtcars$cyl)
# 
# for (cyl_group in names(groups)) {
#   # Print level 4 markdown header
#   cat("#### Cylinders:", cyl_group, "\n\n")
#   
#   # Print the table using knitr::kable
#   print(knitr::kable(head(groups[[cyl_group]]), format = "markdown"))
#   
#   cat("\n\n")  # Spacing between tables
# }
# ```
# 
# :::

```

<br> 

### Level 3 Header

::: {.panel-tabset}

```{r, results='asis'}
# Example using mtcars dataset grouped by cylinder count
groups <- split(mtcars, mtcars$cyl)

for (cyl_group in names(groups)) {
  # Print level 4 markdown header
  cat("#### Cylinders:", cyl_group, "\n\n")
  
  # Print the table using knitr::kable
  print(knitr::kable(head(groups[[cyl_group]]), format = "markdown"))
  
  cat("\n\n")  # Spacing between tables
}
```

:::

<br>

### Applying to ggplot

The only real change from the table version is that you assign the plot to p and call print(p) — the same rule applies as with kable(): inside a loop, ggplot objects must be explicitly printed or knitr drops them silently.

Mixing tables and plots — since both print(knitr::kable(...)) and print(p) work inside the same results='asis' loop, you can easily render a plot and a summary table per tab just by calling both prints before the closing cat("\n\n").


```{r}

# ### Cylinder Groups {.unnumbered}
# 
# ::: {.panel-tabset}
# ```{r, results='asis'}
# library(ggplot2)
# 
# groups <- split(mtcars, mtcars$cyl)
# 
# for (cyl_group in names(groups)) {
#   cat("##### Cylinders:", cyl_group, "\n\n")
#   
#   p <- ggplot(groups[[cyl_group]], aes(x = wt, y = mpg)) +
#     geom_point(colour = "steelblue", size = 3) +
#     geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
#     labs(
#       title = paste("Cylinders:", cyl_group),
#       x     = "Weight (1000 lbs)",
#       y     = "Miles per Gallon"
#     ) +
#     theme_minimal()
#   
#   print(p)
#   
#   cat("\n\n")
# }
# ```
#
# :::

```

#### Cylinder Groups {.unnumbered}

::: {.panel-tabset}

```{r, results='asis'}
library(ggplot2)

groups <- split(mtcars, mtcars$cyl)

for (cyl_group in names(groups)) {
  cat("##### Cylinders:", cyl_group, "\n\n")
  
  p <- ggplot(groups[[cyl_group]], aes(x = wt, y = mpg)) +
    geom_point(colour = "steelblue", size = 3) +
    geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
    labs(
      title = paste("Cylinders:", cyl_group),
      x     = "Weight (1000 lbs)",
      y     = "Miles per Gallon"
    ) +
    theme_minimal()
  
  print(p)
  
  cat("\n\n")
}
```

:::

<br><br>

### HTML style outputs

::: {.panel-tabset}
```{r, results='asis'}
library(flextable)
library(gtsummary)

groups <- split(mtcars, mtcars$cyl)

for (cyl_group in names(groups)) {
  cat("### Cylinders:", cyl_group, "\n\n")
  
  # --- flextable ---
  ft <- flextable(head(groups[[cyl_group]])) |>
    set_caption(paste("Cylinders:", cyl_group)) |>
    autofit()
  
  cat(flextable::flextable_to_rmd(ft), "\n\n")

  # --- OR gtsummary ---
  gt <- groups[[cyl_group]] |>
    select(mpg, wt, hp) |>
    tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd})")) |>
    bold_labels()
  
  cat(as_gt(gt) |> gt::as_raw_html(), "\n\n")
}
```

:::
