---
title: "Tidy Evaluation in R: Data Masking & Non-Standard Evaluation"
format: html
editor: visual
---

## Overview

This document covers **data masking** and **tidy evaluation** in R — core concepts in the `tidyverse`, particularly in `dplyr`, `ggplot2`, and `rlang`. It also flags one factual error found in the original source material.

------------------------------------------------------------------------

## 1. Data Masking

Data masking allows you to refer to columns in a data frame *directly*, without the `df$` prefix.

### Base R (without masking)

```{r}
df <- data.frame(age = sample(1:100, 100, replace = TRUE))
mean(df$age)
```

### With data masking (dplyr)

```{r, collapse=T, message=F, warning=F}
library(tidyverse)

df %>% summarise(mean_age = mean(age))
```

Here, `age` is resolved from within `df` — no `df$` needed.

### Why "masking"?

When `dplyr` evaluates `age`, it first looks *inside* the data frame. If not found there, it falls back to the global environment. This temporary priority is what "masks" the outer scope.

### Packages that use data masking

-   **dplyr**: `filter()`, `mutate()`, `summarise()`
-   **ggplot2**: `aes(x = age, y = score)`
-   **rlang**: provides the underlying tools for programming with masking

------------------------------------------------------------------------

## 2. `enquo()` — Capturing a Single Variable

`enquo()` captures a function argument as a **quosure** — a quoted expression bundled with its environment. This lets you delay evaluation until you're inside a tidy context.

### Without `enquo()` — Broken

```{r, error=T}
my_filter <- function(data, column, value) {
  data %>% filter(column > value)  # column is not found in the data
}

my_filter(df, age, 25)

```

### With `enquo()` — Works

```{r}

my_filter <- function(data, column, value) {
  col_quo <- enquo(column)
  data %>% filter(!!col_quo > value)
}

df <- tibble(age = c(20, 30, 40))
my_filter(df, age, 25)

```

`enquo(column)` captures `age` as a quosure; `!!col_quo` *unquotes* it (injects it) inside `filter()`.

### Related functions

| Function   | Purpose                                    |
|------------|--------------------------------------------|
| `enquo()`  | Capture one argument as a quosure          |
| `enquos()` | Capture multiple arguments as quosures     |
| `quo()`    | Manually create a quosure                  |
| `!!`       | Unquote a single quosure                   |
| `!!!`      | Splice a list of quosures                  |
| `{{ }}`    | Shorthand for `enquo()` + `!!` (preferred) |

------------------------------------------------------------------------

## 3. {{ }} — The Modern Shorthand

{{ var }} combines `enquo()` and `!!` in one clean step. Prefer it for single arguments in most everyday functions.

```{r}
library(dplyr)

group_and_summarise <- function(data, group_col, summary_col) {
  data %>%
    group_by({{ group_col }}) %>%
    summarise(mean_val = mean({{ summary_col }}, na.rm = TRUE), .groups = "drop")
}

df <- tibble(
  gender = c("M", "F", "M", "F"),
  score  = c(85, 90, 88, 95)
)

group_and_summarise(df, gender, score)
```

> **Limitation**: {{ }} handles one variable at a time. For multiple grouping variables, use `enquos()` + `!!!` instead.

------------------------------------------------------------------------

## 4. `enquos()` — Capturing Multiple Variables

`enquos(...)` captures any number of arguments as a **list of quosures**. Use `!!!` to splice them back into tidyverse functions.

```{r}
library(dplyr)
library(rlang)

group_and_summarise_multi <- function(data, ..., summary_var) {
  group_vars   <- enquos(...)
  summary_quo  <- enquo(summary_var)

  data %>%
    group_by(!!!group_vars) %>%
    summarise(mean_val = mean(!!summary_quo, na.rm = TRUE), .groups = "drop")
}

df <- tibble(
  gender = c("M", "F", "M"),
  city   = c("A", "A", "B"),
  score  = c(85, 90, 88)
)

group_and_summarise_multi(df, gender, city, summary_var = score)
```

`enquos()` also works with a single variable, but `enquo()` or `{{ }}` is simpler when you only need one.

------------------------------------------------------------------------

## 5. `!!` and `!!!` Explained

### `!!` — Unquote (single)

Injects one captured quosure into an expression.

```{r}
summarise_one <- function(data, column) {
  col <- enquo(column)
  data %>% summarise(mean_val = mean(!!col))
}
```

### `!!!` — Splice (multiple)

Expands a list of quosures into separate arguments — equivalent to writing them out individually.

```{r}
group_multi <- function(data, ...) {
  group_vars <- enquos(...)

  data %>%
    group_by(!!!group_vars) %>%
    summarise(n = n(), .groups = "drop")
}
```

**Analogy**: if `group_vars` holds `quo(gender)` and `quo(city)`, then `!!!group_vars` inside `group_by()` becomes `group_by(gender, city)`.

------------------------------------------------------------------------

## 6. String-Based Column Selection

When column names are supplied as character strings (common in Shiny or programmatically generated pipelines), use `all_of()` and `.data[[]]`.

```{r}
library(dplyr)

group_and_summarise_str <- function(data, group_cols, summary_col) {
  data %>%
    group_by(across(all_of(group_cols))) %>%
    summarise(mean_val = mean(.data[[summary_col]], na.rm = TRUE), .groups = "drop")
}

df <- tibble(
  gender = c("M", "F", "M", "F"),
  city   = c("A", "A", "B", "B"),
  score  = c(85, 90, 88, 95)
)

group_and_summarise_str(df, group_cols = c("gender", "city"), summary_col = "score")
```

------------------------------------------------------------------------

## 7. `sym()` — Creating Symbols from Strings

`sym()` from `rlang` converts a character string into a symbol that can be unquoted with `!!` inside tidy verbs.

```{r}
library(rlang)
library(dplyr)

my_summarise <- function(data, col_name) {
  sym_col <- sym(col_name)
  data %>%
    summarise(avg = mean(!!sym_col, na.rm = TRUE))
}

df <- tibble(x = c(1, 2, 3))
my_summarise(df, "x")
```

### Comparison of symbol-related tools

| Function         | Purpose                                | Use with           |
|------------------|-------------------------------------|------------------|
| `sym("x")`       | String → symbol (rlang, tidy-friendly) | `!!sym(...)`       |
| `as.symbol("x")` | String → symbol (base R)               | Base R `eval()`    |
| `enquo(x)`       | Capture argument as quosure            | `!!` in tidy verbs |
| `{{ x }}`        | Shorthand capture + unquote            | Inside functions   |

------------------------------------------------------------------------

## Symbols

A symbol (also called a name) is one of R's basic building blocks. 

It represents the name of an object rather than its value. 

When you type age in R, you're using a symbol — R looks it up and returns whatever value is bound to that name. 

A symbol is the thing before the lookup happens.

### Creating symbols explicitly

Normally R creates symbols implicitly, but you can create one directly from a string:

```{r}
as.symbol("x")   # base R — correct
as.name("x")     # base R — also correct, identical to as.symbol()
```

These are base R functions. They're useful when you have a column name as a string and need to turn it into something R can evaluate as code. You can confirm what you get back:

```{r}
x <- as.symbol("age")
typeof(x)   # "symbol"
class(x)    # "name"
```

Note that x now holds the symbol age, not the string "age", and not any value called age. It's a reference waiting to be evaluated.

Why does this matter for tidy evaluation?

Inside dplyr verbs, R expects column references to be symbols, not strings. So if you have a column name stored as a string and want to use it programmatically, you need to convert it first:

```{r, error=T}
df <- data.frame(age = sample(1:100, 100, replace = TRUE))

col_name <- "age"

# This won't work — dplyr sees the string "age", not the column
df %>% summarise(mean(col_name))

# Convert to symbol first, then unquote with !!
col_sym <- as.symbol(col_name)
df %>% summarise(mean(!!col_sym))

```

<br> 

### tidyr symbols 

rlang::sym() — the tidyverse version
rlang::sym() does exactly the same thing as as.symbol(), but is designed to work smoothly within tidy evaluation:

```{r}
rlang::sym("x")  # tidyverse — correct and preferred
```

<br>

```{r, error =T}

col_sym <- sym("age")     # same result as as.symbol("age")
df %>% summarise(mean(!!col_sym, na.rm = TRUE))

```

Prefer sym() in tidyverse code for two reasons. First, it integrates naturally with !! and the rest of the rlang toolkit. Second, there is a plural form syms() that converts a list of strings into a list of symbols, ready to be spliced with !!!:

```{r}

df <- tibble(
  gender = c("M", "F", "M", "F"),
  city   = c("A", "A", "B", "B"),
  score  = c(85, 90, 88, 95)
)

cols <- syms(c("gender", "score", "city"))

df %>% select(!!!cols)

```

This is particularly useful when column names come from a character vector — for example, from a config file, user input, or the result of names().


## 8. Argument Validation: `match.arg()` and `stop()`

When writing functions that accept restricted choices or require
specific inputs, validate arguments early and fail with a clear,
informative error.

### `match.arg()` --- restricting to an allowed set

Declare the default as a vector of options and `match.arg()` handles
both validation and the default (first element) automatically. Partial
matching is also supported.

```{r, error =T}
summarise_col <- function(data, col_name, fn = c("mean", "median", "sum")) {
  fn <- match.arg(fn)

  result <- switch(fn,
    mean   = mean(data[[col_name]],   na.rm = TRUE),
    median = median(data[[col_name]], na.rm = TRUE),
    sum    = sum(data[[col_name]],    na.rm = TRUE)
  )

  tibble(stat = fn, value = result)
}

summarise_col(df, "score")           # uses "mean" by default
summarise_col(df, "score", "median") # explicit choice
summarise_col(df, "score", "med")    # partial matching — also works
summarise_col(df, "score", "non_Stat_value")
```

### `stop()` --- raising custom errors

Use `stop()` for conditions `match.arg()` can't handle, such as checking
that a column exists or is the right type. Place these checks at the top
of the function so the user gets a clear message immediately rather than
a cryptic error from inside a tidy verb.

```{r, error=T}
summarise_col_safe <- function(data, col_name, fn = c("mean", "median", "sum")) {
  if (!col_name %in% names(data)) {
    stop("Column '", col_name, "' not found in data. ",
         "Available columns: ", paste(names(data), collapse = ", "))
  }

  if (!is.numeric(data[[col_name]])) {
    stop("Column '", col_name, "' must be numeric, but is ",
         class(data[[col_name]]), ".")
  }

  fn <- match.arg(fn)
  # ... rest of function
  
  return("met conditions")
  
}

df$city # non numeric
summarise_col_safe(df, "city")

names(df) # Variable name not present
summarise_col_safe(df, "miss_var")

# numeric, non-missing var
df$score
summarise_col_safe(df, "score") 

```

### `stopifnot()` --- compact assertions

For simple yes/no checks, `stopifnot()` is more concise. The named form
(`"message" = expression`) produces readable errors and is available
from R 4.0.0 onwards.

This checks that 

a) variable is a string, and length = 1, 
b) column in data, 
c) column is numeric. 


```{r, error = T}
summarise_col_assert <- function(data, col_name, fn = c("mean", "median", "sum")) {
  stopifnot(
    "col_name must be a single string" = is.character(col_name) && length(col_name) == 1,
    "col_name must exist in data"      = col_name %in% names(data),
    "column must be numeric"           = is.numeric(data[[col_name]])
  )

  fn <- match.arg(fn)
  # ... rest of function
  
  return("Met conditions")
}

df$city # non numeric
summarise_col_assert(df, "city")

names(df) # Variable name not present
summarise_col_assert(df, "miss_var")

# Length > 1
summarise_col_assert(df, c("miss_var", "Score"))

# numeric, non-missing var
df$score
summarise_col_assert(df, "score") 

```

### `rlang::abort()` --- tidyverse-style errors

Preferred over `stop()` when writing functions that integrate closely
with the tidyverse. Produces structured conditions with a catchable
class and better stack traces.

Need look at more.

```{r, error =T}

summarise_col_tidy <- function(data, col_name, fn = c("mean", "median", "sum")) {
  if (!col_name %in% names(data)) {
    abort(
      paste0("Column '", col_name, "' not found in data."),
      class = "col_not_found_error",
      col   = col_name
    )
  }

  fn <- match.arg(fn)
  # ... rest of function
}

# Length > 1
summarise_col_tidy(df, c("miss_var", "score"))


```

The `class` argument lets callers catch specific error types with
`tryCatch()`, which is especially useful in package development or Shiny
apps.

### Summary of validation tools

  -----------------------------------------------------------------------
  Tool                                Best for
  ----------------------------------- -----------------------------------
  `match.arg()`                       Restricting a character arg to a
                                      fixed set of values

  `stop()`                            Custom error messages with flexible
                                      conditions

  `stopifnot()`                       Compact multi-condition assertions
                                      (R 4.0.0+)

  `rlang::abort()`                    Structured, classable errors for
                                      package/tidy code
  -----------------------------------------------------------------------


------------------------------------------------------------------------

## Summary

| Goal                                    | Tool                         |
|-----------------------------------------|------------------------------|
| Use column name in a function (1 var)   | `{{ var }}` or `enquo()`     |
| Use column names in a function (n vars) | `enquos(...)` + `!!!`        |
| Column name as a string                 | `sym()` + `!!`               |
| Multiple string-based column names      | `all_of()` inside `across()` |
| Access column by string in summarise    | `.data[["col"]]`             |
